---
title: "Class 15: Pertussis mini-project"
author: "Samson A16867000"
format: pdf
---

## Background

Pertussis, a.k.a whooping cough, is a highly infectious lunch disease caused by the bacteria *B. Pertussis*

The CDC tracks pertussis cases numbers per year. Lets have a closer look at this data: 

[CDC data ](https://www.cdc.gov/pertussis/php/surveillance/pertussis-cases-by-year.html?CDC_AAref_Val=https://www.cdc.gov/pertussis/surv-reporting/cases-by-year.html)

We will use the **datapasta** R package to "scrape" this data into R. 

```{r}
cdc<-data.frame(
                               year = c(1922L,1923L,1924L,1925L,1926L,
                                        1927L,1928L,1929L,1930L,1931L,1932L,
                                        1933L,1934L,1935L,1936L,1937L,1938L,
                                        1939L,1940L,1941L,1942L,1943L,
                                        1944L,1945L,1946L,1947L,1948L,1949L,
                                        1950L,1951L,1952L,1953L,1954L,1955L,
                                        1956L,1957L,1958L,1959L,1960L,
                                        1961L,1962L,1963L,1964L,1965L,1966L,
                                        1967L,1968L,1969L,1970L,1971L,1972L,
                                        1973L,1974L,1975L,1976L,1977L,1978L,
                                        1979L,1980L,1981L,1982L,1983L,
                                        1984L,1985L,1986L,1987L,1988L,1989L,
                                        1990L,1991L,1992L,1993L,1994L,1995L,
                                        1996L,1997L,1998L,1999L,2000L,
                                        2001L,2002L,2003L,2004L,2005L,2006L,
                                        2007L,2008L,2009L,2010L,2011L,2012L,
                                        2013L,2014L,2015L,2016L,2017L,2018L,
                                        2019L,2020L,2021L,2022L,2024L),
       cases = c(107473,164191,165418,152003,
                                        202210,181411,161799,197371,166914,
                                        172559,215343,179135,265269,180518,
                                        147237,214652,227319,103188,183866,
                                        222202,191383,191890,109873,133792,
                                        109860,156517,74715,69479,120718,68687,
                                        45030,37129,60886,62786,31732,28295,
                                        32148,40005,14809,11468,17749,
                                        17135,13005,6799,7717,9718,4810,3285,
                                        4249,3036,3287,1759,2402,1738,
                                        1010,2177,2063,1623,1730,1248,1895,
                                        2463,2276,3589,4195,2823,3450,4157,
                                        4570,2719,4083,6586,4617,5137,
                                        7796,6564,7405,7298,7867,7580,9771,
                                        11647,25827,25616,15632,10454,13278,
                                        16858,27550,18719,48277,28639,
                                        32971,20762,17972,18975,15609,18617,
                                        6124,2116,3044,23544)
     )

```

>. Q1. With the help of the R “addin” package datapasta assign the CDC pertussis case number data to a data frame called cdc and use ggplot to make a plot of cases numbers over time.

```{r}
library(ggplot2)

baseplot <- ggplot(cdc) +
  aes(year, cases)+
  geom_point() +
  geom_line()

baseplot
```

>. Q2. Using the ggplot geom_vline() function add lines to your previous plot for the 1946 introduction of the wP vaccine and the 1996 switch to aP vaccine (see example in the hint below). What do you notice?

```{r}
baseplot + 
  geom_vline(xintercept = 1946, col = "red") +
  geom_vline(xintercept = 1996, col = "blue") +
  geom_vline(xintercept = 2020, col = "gray") + 
  geom_vline(xintercept = 2003, col = "green")
```

We went from ~200,000 cases pre wP vaccine to ~1,000 cases in 1976. The US switched to the aP vaccine in 1995. We start to see a big increase in 2004 to ~26,000 cases. 

>. Q3. Describe what happened after the introduction of the aP vaccine? Do you have a possible explanation for the observed trend?

After a few years of the itnroduction of the aP vaccine we see a big increase to 26,000 cases. People don't want to get vaccinated (vaccination rates are lower than before). There can also be different variants leading to more cases. 

There is a ~10 year lag from aP roll out to increasing case numbers. This holds true of other countries like Japan, UK, etc.

**Key question**: why does the aP vaccine induced immunity wane faster than that of the wP vaccine? 

## CMI-PB 

The CMI-PB(Computational Models of Immunity Pertussis Boost) makes available lots of data about the immune response to Pertussis booster vaccination.

Critically, it tracks wP and aP individuals over time to see how their immune response changes. 

CMI-PB make all their data freely available via KSON format tables from their database.

Let's read the first one of these tables: 

```{r}
library(jsonlite)

subject <- read_json("http://cmi-pb.org/api/v5/subject",
                     simplifyVector = TRUE)
head(subject)
```

>. Q. How many subjects are there in this dataset?

```{r}
nrow(subject)
```

>. Q4. How many aP and wP infancy vaccinated subjects are in the dataset?

```{r}
table(subject$infancy_vac)
```
>. Q5. How many Male and Female subjects/patients are in the dataset?

```{r}
table(subject$biological_sex)
```

>. Q6. What is the breakdown of race and biological sex (e.g. number of Asian females, White males etc…)?

```{r}
table(subject$race, subject$biological_sex)
```

>. Q Does this do a good job of representing the US populus?

No, because we can see a small representation of American Indian, Native Hawaiian, Black individuals. 

```{r}
specimen <- read_json("http://cmi-pb.org/api/v5/specimen", 
                      simplifyVector = TRUE)
head(specimen)
```

Now we can join these two tables `subkect` and `specimen` to make one new `meta` table with the combined data.

>. Q9. Complete the code to join specimen and subject tables to make a new merged data frame containing all specimen records along with their associated subject details:

```{r}
library(dplyr)

meta <- inner_join(specimen, subject)
dim(meta)
head(meta)
```

Now read an "experiment data" table from CMI-PB

>. Q10. Now using the same procedure join meta with titer data so we can further analyze this data in terms of time of visit aP/wP, male/female etc.


```{r}
abdata <- read_json("http://cmi-pb.org/api/v5/plasma_ab_titer", 
                      simplifyVector = TRUE)
head(abdata)
```

One more join to do of `meta` and `abdata` to associate all the meta data about the individual and their race, biological sex and infancy vaccination status together with Antibody levels...


```{r}
ab <- inner_join(abdata, meta)
head(ab)
```

```{r}
nrow(ab)
```
>. Q11. How many specimens (i.e. entries in abdata) do we have for each isotype?

```{r}
table(ab$isotype)
```

```{r}
table(ab$antigen)
```

>. Q12. What are the different $dataset values in abdata and what do you notice about the number of rows for the most “recent” dataset?

```{r}
table(ab$dataset)
```
The different $dataset values shows that the number of rows for the most "recent" data is the fewest compared to years past. 


Let's focus in on IgG - one of the main antibody types responsive to bacteria or viral infections

```{r}
igg <- filter(ab, isotype == "IgG")
head(igg)
```

Make a first plot of MFI (Mean Flurosence Intensity - measure of how is detected) for each antigen. 

>. Q13. Complete the following code to make a summary boxplot of Ab titer levels (MFI) for all antigens:


```{r}
ggplot(igg) +
  aes(MFI_normalised, antigen) +
  geom_boxplot() +
  facet_wrap(vars(visit), nrow=2)
```

```{r}
ggplot(igg) +
  aes(MFI_normalised, antigen, col=infancy_vac ) +
  geom_boxplot() +
  facet_wrap(vars(visit), nrow=2)
```

```{r}
table(igg$visit)
```


```{r}
igg_7 <- filter(igg, visit %in% 1:7)
table(igg_7$visit)
```


```{r}
ggplot(igg_7) +
  aes(MFI_normalised, antigen) +
  geom_boxplot() +
  facet_wrap(~visit)
```


```{r}
ggplot(igg_7) +
  aes(MFI_normalised, antigen, col=infancy_vac ) +
  geom_boxplot(show.legend = FALSE) + 
  facet_wrap(vars(visit), nrow=2) +
  xlim(0,75) +
  theme_bw()
```

>. Q14. What antigens show differences in the level of IgG antibody titers recognizing them over time? Why these and not others?

An antigen that shows differences in level of IgG antibody titers is the "PT"

```{r}
igg_7 %>% filter(visit != 8) %>%
ggplot() +
  aes(MFI_normalised, antigen, col=infancy_vac ) +
  geom_boxplot(show.legend = FALSE) + 
  xlim(0,75) +
  facet_wrap(vars(infancy_vac, visit), nrow=2)
```

```{r}
abdata.21 <- ab %>% filter(dataset == "2021_dataset")

abdata.21 %>% 
  filter(isotype == "IgG",  antigen == "PT") %>%
  ggplot() +
    aes(x=planned_day_relative_to_boost,
        y=MFI_normalised,
        col=infancy_vac,
        group=subject_id) +
    geom_point() +
    geom_line() 
```



