---
title: "Class 10: Halloween Mini Project"
author: "Samson A16867000"
format: pdf
---

---
title: "Class 10: Halloween Mini Project"
author: "Samson A16867000"
format: pdf
---

Today is Halloween, an ole Irish holiday, let's celebrate by eating candy. 

We will explore some data all about Halloween candy from the 538 website. 

```{r}
candy_file <- "https://raw.githubusercontent.com/fivethirtyeight/data/master/candy-power-ranking/candy-data.csv"

candy = read.csv(candy_file, row.names = 1)
head(candy)
```

>. Q1. How many different candy types are in this dataset?

```{r}
nrow(candy)
```
```{r}
rownames(candy)
```


>. Q2. How many fruity candy types are in the dataset?

```{r}
sum(candy$fruity)
```
```{r}
candy["Twix", ]$winpercent
```
>. Q3. What is your favorite candy in the dataset and what is it’s winpercent value?

```{r}
candy["Milky Way", ]$winpercent
```

>. Q4. What is the winpercent value for “Kit Kat”?

```{r}
candy["Kit Kat", ]$winpercent
```

>. Q5. What is the winpercent value for “Tootsie Roll Snack Bars”?              

```{r}
candy["Tootsie Roll Snack Bars", ]$winpercent
```
```{r}
library(dplyr)

candy |>
    filter(rownames(candy)== "Haribo Happy Cola") |>
    select(winpercent)
```

>. Q. FInd fruity candy with a winpercent above 50 

```{r}
candy |> 
    filter(winpercent >50) |> 
    filter(fruity ==1)
```

```{r}
top.candy <- candy[candy$winpercent >50,]
top.candy[top.candy$fruity ==1,]
```

To get q quikc insihgt into a new data set some folks like using the skier package and its `skim` function

```{r}
skimr::skim(candy)
```

Looks like the `winpercent` vatriable or column is measured on a different scale than everything else!I will need to scale my data ebefore doing any analysis like PCA etc. 

>. Q6. Is there any variable/column that looks to be on a different scale to the majority of the other columns in the dataset?

The variable that looks to be on a different scale is winpercent

>. Q7. What do you think a zero and one represent for the candy$chocolate column?

zero represents the candy not being chocolate and one represents the candy being chocolate 

>. Q8. Plot a histogram of winpercent values

We can do this a few ways, e.g the "base" R `hist()` function or with `ggplot`

```{r}
hist(candy$winpercent, breaks = 30)
```

library(ggplot2)

```{r}
library(ggplot2)

ggplot(candy) +
  aes(winpercent) +
  geom_histogram(binwidth = 8) +
  theme_bw()
```

ggplot(candy) +
  aes)winpercent +
  geom_histogram(binwidth = 8) +
  theme_bw
>. Q9. Is the distribution of winpercent values symmetrical?

The distributionof winpercent values are not symmetrical

>. Q10. Is the center of the distribution above or below 50%?

```{r}
summary(candy$winpercent)
```

The center of the distribution is below 50% when taking into account the median

>. Q11. On average is chocolate candy higher or lower ranked than fruit candy?

```{r}
fruit.candy <- candy |>
    filter(fruity ==1) 
  
summary(fruit.candy$winpercent)
```
```{r}
chocolate.candy <- candy |>
    filter(chocolate ==1) 
  
summary(chocolate.candy$winpercent)
```

On average chocolate is higher ranked than fruit candy 

>. Q12. Is this difference statistically significant?

```{r}
t.test(chocolate.candy$winpercent, fruit.candy$winpercent)
```
The difference between chocolate and fruity candy is statistically significant 

```{r}
play <- c("d","a","c")
sort(play)
order

play[order(play)]
```

>. Q13. What are the five least liked candy types in this set?

```{r}
head(candy[order(candy$winpercent),],5)
```
>. Q14. What are the top 5 all time favorite candy types out of this set?

```{r}
tail(candy[order(candy$winpercent),],5)
```
>. Q15. Make a first barplot of candy ranking based on winpercent values.

```{r}
ggplot(candy) + 
  aes(winpercent,rownames(candy)) +
  geom_col()
```

>. Q16. This is quite ugly, use the reorder() function to get the bars sorted by winpercent?

```{r}
ggplot(candy) + 
  aes(winpercent, 
      reorder(rownames(candy),winpercent),) +
  geom_col()
```



```{r}
ggplot(candy) + 
  aes(winpercent, 
      reorder(rownames(candy),winpercent),
      fill = chocolate) +
  geom_col()
```

I want a more custome color scheme where I can see both chocolate and bar and fruity etc. all from the one plot. To do this we can roll our own color vector...

```{r}
# Place holder color vector
mycols <- rep("black", nrow(candy))
mycols[as.logical(candy$chocolate)] <- "chocolate"
mycols[as.logical(candy$bar)] <- "brown"
mycols[as.logical(candy$fruity)] <- "pink"
mycols[rownames(candy) =="Milky Way"] <- "blue"

mycols
```


```{r}
ggplot(candy) + 
  aes(winpercent, 
      reorder(rownames(candy),winpercent),
      fill = chocolate) +
  geom_col(fill = mycols)
```

>. - Q17. What is the worst ranked chocolate candy?

Sixlets 

>. Q18. What is the best ranked fruity candy?

Starbursts

PLot of winpercent vs pricepercent to see what would be the best candy to buy... 


```{r}
mycols[as.logical(candy$fruity)] <- "green"

```


```{r}
library(ggrepel)


ggplot(candy) +
  aes(winpercent, pricepercent, label=rownames(candy)) +
  geom_point(col=mycols) + 
  geom_text_repel(col=mycols, size=3.3, max.overlaps = 8)
```

>. Q19. Which candy type is the highest ranked in terms of winpercent for the least money - i.e. offers the most bang for your buck?

Reese's Miniatures

>. Q20. What are the top 5 most expensive candy types in the dataset and of these which is the least popular?

```{r}
tail(candy[order(candy$pricepercent),],5)
```
The least popular out of the 5 most expensive candy types is Nik L Nip

```{r}
library(corrplot)
```

```{r}
cij <- cor(candy)
corrplot(cij, diag = F)
```

>. Q22. Examining this plot what two variables are anti-correlated (i.e. have minus values)?

Chocolate and fruity

>. Q23. Similarly, what two variables are most positively correlated?

Chocolate and bar

# Principle Component Analysis

```{r}
pca <- prcomp(candy, scale = TRUE)
summary(pca)
```

```{r}
plot(pca$x[,1],pca$x[,2], col = mycols, pch =16)
```

How do the original variables (columns) contribute to the new PCs. I will look at PC1 here. 

```{r}
loadings <- as.data.frame(pca$rotation)

ggplot(loadings)+
    aes(PC1,reorder(rownames(loadings),PC1), fill=PC1)+
    geom_col()
```

>. Q24. What original variables are picked up strongly by PC1 in the positive direction? Do these make sense to you?

fruity, pluribus, and hard. It makes sense because it is representative of 
what we saw from the correlation table when we used `corrplot` 

```{r}
pc.results <- cbind(candy, pca$x)

ggplot(pc.results, aes(x = PC1, y = PC2, label = rownames(pc.results))) +
  geom_point(col = mycols) +
  geom_text_repel(col = mycols) +
  labs(title = "Candy Space via PCA")
```

