---
title: "HW Class 12"
author: "Samson A16867000"
date: "2024-11-09"
output: pdf
---

## Section 4: Population Scale Analysis

>. Q13: Read this file into R and determine the sample size for each genotype and their
corresponding median expression levels for each of these genotypes.


```{r}
expr <- read.table("rs8067378_ENSG00000172057.6.txt")
head(expr)
```

```{r}
nrow(expr)
```

```{r}
table(expr$geno)
```

```{r}
geno.exp <- boxplot(exp ~ geno, data = expr)
```
```{r}
medians <- aggregate(exp ~ geno, data = expr, FUN = median)
medians
```

median expression levels for each of the genotypes


```{r}
library(ggplot2)
```

>. Q14: Generate a boxplot with a box per genotype, what could you infer from the relative
expression value between A/A and G/G displayed in this plot? Does the SNP effect the
expression of ORMDL3?

```{r}
ggplot(expr) + aes(geno, exp, fill = geno)+
  geom_boxplot(notch=TRUE)
```
Based on the boxplot it seems that having G/G is associated with having less expression compared to A/A. Yes the SNP effects the expression of ORMDL3
